---
title: "Compositional analysis of microbiome data"
author: "Xin (David) Zhao"
date: "Last edited `r format(Sys.time(), '%d %B %Y')`"
knit: (function(inputFile, encoding) {
      out_dir <- 'docs';
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_file=file.path(dirname(inputFile), out_dir, 'index.html'))})
output:
  html_document:
    theme: cosmo
    highlight: pygments
    df_print: paged
    toc: TRUE
    toc_float: TRUE
    collapsed: FALSE
    number_sections: TRUE
    fig_width: 7
    fig_height: 6
    fig_caption: TRUE
--- 

```{r setup, include = FALSE}
# set options for the entire document 
knitr::opts_chunk$set(fig.align = 'center', 
                      fig.dim =c(6,4), 
                      dev="png",
                      echo=TRUE, #display code in output document 
                      error=FALSE,
                      message=FALSE) #stop render when error occurs   
```


This project was built with R `r getRversion()`.
```{r load libraries, message=FALSE, warning=FALSE}

library(ALDEx2)
library(vegan)
library(xtable)  #make result table
library(zCompositions) # replace 0 values 
library(tidyverse)

```

## Introduction 
- Compositional data 
- Aitchison simplex 
- A family of log-ratio transformation 
- How to deal zeros in compositional data analysis
- Statistical tools for compositional data analysis



## Outline of steps to compare between groups using ALDEx2 package 
1. Download raw microbiome data and metadata 
2. Preprocess input data - replace 0 values using the `zCompositions` package 
3. Convert data to proportions 
4. Perform abundance and sample filtering and deal sparsity 
5. Perform the `clr` data transform 
6. Perform the singular value decomposition using the function `prcomp()` 
7. Display the results of PCoA by using `biplot()` or `coloredBiplot()` 
8. Visualization for exploratory data analysis 
9. Perform the Welch's t and Wilcoxon Rank Sum Test using `aldex.ttest()` 
10. Estimate effect size using the function `aldex.effect()` 
11. Merge all data into one object and make a data frame for result viewing and downstream analysis 
12. Difference plot, effect size and effect plots 

Steps through 1 to 8 are data exploratory while steps through 9 to 12 are differential abundance analysis by `ALDEx2` 


## Scripts to implement above procedures 

### Step 1. Download raw microbiome data and metadata  
The microbiome OTU table and metadata was retrieved from [ML Repo] (https://knights-lab.github.io/MLRepo/docs/turnbaugh_lean_obese_all.html)  

```{r}

# raw OTU table
raw_otu <- read.csv(file = "https://knights-lab.github.io/MLRepo/datasets/turnbaugh/refseq/otutable.txt",
                    header=T,
                    sep = "")

glimpse(raw_otu) 

```
Read in the metadata from URL. 

```{r}

# metadata 

meta <- read.csv(file = "https://knights-lab.github.io/MLRepo/datasets/turnbaugh/task-obese-lean-all.txt",
         header = TRUE,
         sep = "") 

glimpse(meta)

```
The metadata contains `r dim(meta)[1]` samples. `Var` indicates the independent, binary variable of interest, `Lean` and `Obese`. 


### Step 2. Preprocess input data - replace 0 values using the `zCompositions` package  

Convert the raw OTU table, with samples being rows while OTU being columns. 
```{r}

t_raw_otu <- raw_otu %>% 
  gather(key = "sample",
         value = "reads",
         3:ncol(.)) %>% 
  dplyr::select(-ID) %>% 
  spread(key = names(.)[1],
         value = "reads") %>% 
  column_to_rownames("sample")  # convert column to row id 
  
glimpse(t_raw_otu)

```


Replace 0 values using the `zCompositions` package. 

Inspect whether any NA (missing values) exits; replace NA with 0 values if any. Follow the instructions introduced [here] (https://www.r-bloggers.com/2019/09/handling-missing-values-in-r-using-tidyr/). 

```{r}

sum(is.na(t_raw_otu)) # check number of missing values 

na_t_raw_otu <- t_raw_otu %>% 
  mutate_all(replace_na, 0) # replace NA with 0 value 

sum(is.na(na_t_raw_otu)) # check number of missing values afterwards 


```

Evaluate 0/missing values patterns in the OTU table with `zPatterns()` from the `zCompositions` package. 
```{r zPatterns}

pattern.ID <- zPatterns(na_t_raw_otu, 
          label = 0, # identify zero percentages 
          plot=TRUE,
          show.means = FALSE,
          bar.ordered = c(TRUE, TRUE),
          bar.labels = TRUE)  

```


Apply `cmultRepl()` from the `zCompositions` package to replace 0 values with the count zero multiplicative method and output counts. Error information appear and the function stops. Therefore, remove samples that are zeros for every OTU. And then implement the above function again. 
```{r drop OTU with all zeros, collapse=TRUE} 

rm_na_t_raw_otu <- na_t_raw_otu[pattern.ID != 281, ] 


rp_rm_na_t_raw_otu <- cmultRepl(rm_na_t_raw_otu, 
          method = "CZM",
          output = "p-counts",  # pseudo-counts 
          z.warning=0.8)  

```


### Step 3. Convert data to proportions  

Transpose the resulting data frame to OTU X sample format. 

```{r} 

pcount_otu <- rp_rm_na_t_raw_otu %>% 
  rownames_to_column("sample") %>% 
  gather(key = "otu", value = "reads", -sample) %>% 
  spread(key = "sample", value = "reads") %>% 
  column_to_rownames("otu") 

head(pcount_otu)
```

Calculate total pcount throughout samples. 
```{r prop}

pcount_otu %>% 
  rownames_to_column("otu") %>% 
  gather(key= "sample", value = "pcount", -otu) %>% 
  group_by(sample) %>% 
  summarise(total = sum(pcount), .groups = "drop") 

```


Calculate proportions based on pseudo counts.  
```{r, collapse=TRUE}

prop_pcount_otu <- pcount_otu %>% 
  map_dfr(function(x) x/sum(x)) 

colSums(prop_pcount_otu) == 1 # check if sum to each column equals 1 
  
rownames(prop_pcount_otu) <- rownames(pcount_otu)  # assign row names to the resulting data frame 

```

### Step 4. Perform abundance and sample filtering and deal sparsity 

Filter proportional OTU data to remove all OTU that are less than 0.01% in any samples. 

```{r filtering}

otu_id_dropped <- prop_pcount_otu %>% 
  rownames_to_column("otu") %>% 
  gather(key = "sample", value="prop", -otu) %>% 
  group_by(otu) %>%
  summarise(max = max(prop)) %>% 
  ungroup() %>% 
  filter(max < 0.0001) %>% # minimal proportional threshold 0.01% 
  pull(otu)
  
# remove OTU not passing 0.01% threshold 
pcount_otu_fl <- pcount_otu %>% 
  rownames_to_column("otu") %>% 
  filter(!otu %in% otu_id_dropped) %>% 
  column_to_rownames("otu")

```


### Step 5. Perform the clr data transformation

The formula of `clr` is ... 

```{r}

# extract the descendingly ordered taxa 
desc_otu_id <- pcount_otu_fl %>% 
  rownames_to_column("otu") %>% 
  gather(key = "sample", value = "pcount", -otu) %>% 
  group_by(otu) %>% 
  summarise(total = sum(pcount)) %>% 
  arrange(desc(total)) %>% 
  pull(otu)


# re-order the pcount OTU table 
pcount_otu_fl_desc <- pcount_otu_fl[desc_otu_id, ]


# clr transformation 
pcount_otu_fl_desc_clr <- t(apply(pcount_otu_fl_desc, 2, function(x) {log(x) - mean(log(x))}))  

pcount_otu_fl_desc_clr <- data.frame(pcount_otu_fl_desc_clr)  # convert to data frame format 


```

















